Deep Learning does not Replace Bayesian Modeling

### 2021-11-15 DeepLearningAndBayesianSoftware.Rmd
try 2 at final

---
title: "Deep Learning does not Replace Bayesian Modeling"
subtitle: "Comparing research use via citation counting"
author: "Breck Baldwin"
email: "breckbaldwin@gmail.com"
date: "11/01/2021"
output:
 bookdown::html_document2:
   includes:
     in_header: _html/ga.html

@@

@@

Another case are hybrid models where a deep learning component replaces some or all of the likelihood of a Bayesian model. While the overall model may be interpretable, the individual deep learning components remain black boxes. Interpretability for the overall model comes from have well understood properties of the deep learning components, e.g., how they were trained etc. For example Maggie Lieu's presentation at StanCon that used a deep learning component to substitute for a numerically unstable halo mass function (@lieuAstro).

@@

In the end I believe model interpretability cleanly differentiates the the deep learners and Bayesian modelers[^3], at least in practice.

@@

In the process of grant writing I create metrics to help justify projects, lately in partnership with PyMC and ArviZ through the scientific fiscal sponsor NumFOCUS of which Stan is a member as well. Since Bayesian modeling seems always in the shadow of deep learning I started tracking deep learning software packages as well for comparison. Below we see the relative citations of the top deep learning packages TensorFlow, PyTorch and the support package Keras to the Bayesian packages PyMC3, Stan with support/derivative packages RStan, RStanArm, PyStan and brms.

@@

plot2 <-
  ggplot(data = df_long_label, aes(
    x = years,
    y = yr_count,
    group = topic,
    color = topic
  )) +
  geom_line() +
  geom_point() +
  geom_label_repel(aes(label = label),
                   na.rm = TRUE) +
  scale_color_discrete(guide = FALSE) +
  labs(y = "year count for research articles matching topic",
  caption = "Fig 1: Relative counts with and without computer science journals") +
   theme(plot.caption=element_text(size=12, hjust=0, margin=margin(15,0,0,0)))

   log_plot = plot2 + 
     scale_y_continuous(breaks=c(1, 10, 100, 1000, 10000, 100000, 1e+06, 1e+07, 1e+08), 
                     trans = scales::log_trans())
print(plot2)
```

@@

Below are the queries used to identify the packages. Direct citation linking does not work well with these packages since many of them had no journal publication to cite in a referring article. As a result a string search was used to identify the citations in the reference section which was determined by Scopus.com's document parser. I would estimate a 5% false positive rate based based on informal examination of return sets.

@@

The above queries can be run with the advance search option at https://scopus.com and are generated from the API queries actually used where '+' was replaced with ' ' above. An account is required which is included for many education institutions.

@@

# Access to source/scripts and acknowledgements

@@

I'd like to thank Andrew Gelman and the Laplace lab at Columbia University for support.

>>>>

---
title: "Deep Learning does not Replace Bayesian Modeling"
subtitle: "Comparing research use via citation counting"
author: "Breck Baldwin, Columbia University"
email: "breckbaldwin@gmail.com"
affiliation: Columbia Univeristy

output:
 bookdown::html_document2:
   includes:
     in_header: _html/ga.html

@@

<!--

    A short informative title containing the major key words. The title should not contain abbreviations (see Wiley's best practice SEO tips);
    A short running title of less than 40 characters;
    The full names of the authors;
    The author's institutional affiliations where the work was conducted, with a footnote for the author’s present address if different from where the work was conducted;
    Acknowledgements;
    Abstract and keywords;
    Main text;
    References;
    Tables (each table complete with title and footnotes);
    Figure legends;
    Appendices (if relevant).
-->

@@

Another case includes hybrid models where a deep learning component replaces some or all of the likelihood of a Bayesian model. While the overall model may be interpretable, the individual deep learning components remain black boxes. Interpretability for the overall model comes via having well understood properties for the deep learning components, e.g., how they were trained etc. or other methods of characterizing their role. For example Maggie Lieu's presentation at StanCon used a deep learning component to substitute for a numerically unstable halo mass function (@lieuAstro).

@@

In the end I believe model interpretability cleanly differentiates the deep learners and Bayesian modelers[^3], at least in practice.

@@

In the process of grant writing I create metrics to help justify projects, lately in partnership with PyMC and ArviZ through the scientific fiscal sponsor NumFOCUS of which Stan is a member as well. Since Bayesian modeling seems always in the shadow of deep learning I started tracking deep learning software packages as well for comparison. Below we see the relative citations of the top deep learning packages TensorFlow, PyTorch and the support package Keras to the Bayesian packages PyMC3, Stan with supporting/derivative packages RStan, RStanArm, PyStan and brms.

@@

plot2 <-
  ggplot(data = df_long_label, aes(
    x = years,
    y = yr_count,
    group = topic,
    color = topic
  )) +
  geom_line() +
  geom_point() +
  geom_label_repel(aes(label = label),
                   na.rm = TRUE) +
  scale_color_discrete(guide = FALSE) +
  labs(y = "Year count for research articles matching topic",
  caption = "Fig 1: Relative counts with and without computer science journals") +
   theme(plot.caption=element_text(size=12, hjust=0, margin=margin(15,0,0,0)))

   log_plot = plot2 + 
     scale_y_continuous(breaks=c(1, 10, 100, 1000, 10000, 100000, 1e+06, 1e+07, 1e+08), 
                     trans = scales::log_trans())
ggsave('fig1.png', plot=plot2)
print(plot2)
```

@@

Below are the queries used to identify the packages. Direct citation linking does not work well with these packages since many of them had no journal publication to cite in a referring article. Consequently a string search was used to identify the citations in the reference section which was determined by Scopus.com's document parser. I would estimate a 5% false positive rate based based on informal examination of return sets.

@@

The above queries can be run with the advance search option at https://scopus.com and are generated from the API queries actually used where '+' is replaced with ' ' above. An account is required which is included for many education institutions.

@@

# Access to source/scripts

@@

# Acknowledgements

I'd like to thank Andrew Gelman and the Laplace lab at Columbia University. The research was supported by internal Columbia funds.
